Unbiased pangenome graphs

### 2022-02-07 manuscript/main.tex
add cost of bitvector to mark transclosure progress

%Although the space bounds alone indicate that this approach is infeasible, we can also consider its time complexity.
%First, we must enable random access to match elements in $A$.
%Assuming that $A$ contains pairs of characters in $Q$, then we can sort it
For the sake of time and space complexity analysis, we consider a simple algorithm to implement the induction process.
The induction depends on our ability to compute transitive closures of matches $m^+$.
If $A$ is sorted, we can find the matches of a given character in $Q$ using binary search, which allows us to compute $m^+$ for each character.
We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure (e.g. a bitvector of $|Q|$).
As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the sequence-to-graph mapping $Z$, which, like $A$, consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
As with $A$, we can sort $Z$ to obtain random access via binary search.
Finally, we derive elements in $\mathcal{P}$ by iterating through the characters of $Q$ and looking up their mapping in $Z$ using binary search.
The edge set $\mathcal{E}$ are the unique pairs of steps found in $\mathcal{P}$, and can be computed by sorting pairs of steps in $\mathcal{P}$.

@@

The inductive proof of theorem \ref{thm:seqwish} demonstrates how to build a variation graph from sequences and their pairwise alignments.
However, a na\"{i}ve algorithm based on this model would require a very large amount of space.
Take $\mathcal{Q} = |S|$.
Although our identifier space $\mathcal{Q}$ must include all of $Q$, in practice, we only store $S$, as $\bar{S}$ can be trivially computed.
Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $\mathcal{Q}/N$ nodes.
The induction must maintain reference to all characters in all input sequences $O(\mathcal{Q})$, all character-to-character matches $O(|A|) \approx O(\mathcal{Q}N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(\mathcal{Q})$, the nodes of the graph $O(|\mathcal{N}|) \approx O(\mathcal{Q}/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(\mathcal{Q})$, and the set of paths $O(|\mathcal{P}|) \approx O(\mathcal{Q})$.
In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(\mathcal{Q}N^2 + 4\mathcal{Q} + \mathcal{Q}/N)$ space.

Assuming that we want to build a graph, e.g., 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx O(\num{e15})$ identifiers to store the full model.
Such a design is almost infeasible for inputs larger than a handful of genomes.
For instance, we would need $O(\num{3e12})$ identifiers for just 5 human genomes.
Although it is feasible to compute such a graph using external memory, the approximately 200-fold increase in space relative to the input renders this clearly impractical.

@@

\begin{table*}[!ht]
    \centering
    \caption{Performance of the graph induction algorithm.
        For each pangenome we have chosen a single experiment with the given \textit{seqwish -k} filter setting of 49bp.
        }
    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
        \hline
        species & sequences & haplotypes & fasta.Gbps & min.match.bp & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbps & components \\ \hline
        \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline
        \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 39723 & 347.9261 & 464.3392 & 6.2322 & 724 \\ \hline
        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 475 & 33.7270 & 11.9739 & 0.0302351 & 33 \\ \hline
        \textit{Z. mays} & 46289 & 41 & 90.2491 & 49 & 31043 & 351.1235 & 402.8716 & 13.8838 & 925 \\ \hline
    \end{tabular}
\end{table*}

>>>>

%Although the space bounds alone indicate that this approach is infeasible, we can also consider its time complexity.
%First, we must enable random access to match elements in $A$.
%Assuming that $A$ contains pairs of characters in $Q$, then we can sort it
For the sake of time and space complexity analysis, we consider a simple algorithm to implement the induction process.
The induction depends on our ability to compute transitive closures of matches $m^+$.
If $A$ is sorted, we can find the matches of a given character in $Q$ using binary search, which allows us to compute $m^+$ for each character.
We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure $\mathcal{X}$, which could be encoded as a bitvector of length $|Q|$.
As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the sequence-to-graph mapping $Z$, which, like $A$, consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
As with $A$, we can sort $Z$ to obtain random access via binary search.
Finally, we derive elements in $\mathcal{P}$ by iterating through the characters of $Q$ and looking up their mapping in $Z$ using binary search.
The edge set $\mathcal{E}$ are the unique pairs of steps found in $\mathcal{P}$, and can be computed by sorting pairs of steps in $\mathcal{P}$.

@@

The inductive proof of theorem \ref{thm:seqwish} demonstrates how to build a variation graph from sequences and their pairwise alignments.
However, a na\"{i}ve algorithm based on this model would require a very large amount of space.
Take $\mathcal{Q} = |S|$.
Although our identifier space $\mathcal{Q}$ must include all of $Q$, in practice, we only store $S$, as $\bar{S}$ can be trivially computed.
Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $\mathcal{Q}/N$ nodes.
The induction must maintain reference to all characters in all input sequences $O(\mathcal{Q})$, all character-to-character matches $O(|A|) \approx O(\mathcal{Q}N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(\mathcal{Q})$, the nodes of the graph $O(|\mathcal{N}|) \approx O(\mathcal{Q}/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(\mathcal{Q})$, and the set of paths $O(|\mathcal{P}|) \approx O(\mathcal{Q})$.
We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(\mathcal{Q})$ bits, equivalent to $O(\mathcal{Q} / \log_2\mathcal{Q})$ integer identifiers.
In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(\mathcal{Q}(N^2 + 1/N + 1/log_2{Q} + 4))$ space.

Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
Such a design is almost infeasible for inputs larger than a handful of genomes.
For instance, we would need $\approx \num{3e12}$ identifiers for just 5 human genomes.
Although it is feasible to compute such a graph using external memory, the approximately 200-fold increase in space relative to the input renders this clearly impractical.

@@

\begin{table*}[!ht]
    \centering
    \caption{Performance of the graph induction algorithm.
        For each pangenome we have chosen a single experiment with the given \textit{seqwish -k} filter setting of 49bp.
        }
    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
        \hline
        species & sequences & haplotypes & fasta.Gbps & min.match.bp & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbps & components \\ \hline
        \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline
        \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 46268 & 347.4983 & 604.4261 & 4.47126 & 474 \\ \hline
        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 790 & 75.7514 & 20.2070 & 0.01421 & 5 \\ \hline
        \textit{Z. mays} & 46289 & 41 & 90.2491 & 49 & 31043 & 351.1235 & 402.8716 & 13.8838 & 925 \\ \hline
    \end{tabular}
\end{table*}
